### 2018-08-29 doc/large_strain.tex

%\documentclass[12pt]{article}
\documentclass[preprint,12pt,times]{elsarticle}

@@

% \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
%\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{bm}
%\usepackage{bbm}
%\usepackage{caption}
\usepackage{color}
%\usepackage{geometry}
%\usepackage{graphicx}
%\usepackage[colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
%\usepackage{layouts}
%\usepackage{mathrsfs}
%\usepackage{pgfkeys,pgfmath,pgfcore}
%\usepackage{srcltx}
%\usepackage[labelformat=simple]{subcaption}
%\usepackage{xspace}
\usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}

@@

\section{Matrix-free operator evaluation}

{\color{red}
[DA] Daniel, please add a brief intro about MF (readership will likely not know anything about it), maybe borrow 1-2 pictures to explain how tensor-product structure is handled.
}

@@

TODO: question to asnwer -- how much to cache? At least $detF ^ {2/3}$. Other options is stress and strain at quadrature points.
Run benchmarks and see.

@@

TODO: add new results from Cook's and some matrix-inclusion (matrix-pore).
Eliptical holes are easier as we don't need heterogeneous operator.

>>>>

\RequirePackage[l2tabu, orthodox]{nag}
%\documentclass[12pt]{article}
\documentclass[preprint,12pt,times]{elsarticle}

@@

% \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
%\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{bm}
%\usepackage{bbm}
%\usepackage{caption}
\usepackage{xcolor}
%\usepackage{geometry}
%\usepackage{graphicx}
%\usepackage[colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{listings}
\lstset{ %
   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
   basicstyle=\small,        % the size of the fonts that are used for the code
   breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
   breaklines=false,                 % sets automatic line breaking
   captionpos=b,                    % sets the caption-position to bottom
   %commentstyle=\color{mygreen},    % comment style
   deletekeywords={...},            % if you want to delete keywords from the given language
   escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
   extendedchars=false,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
   frame=single,	                   % adds a frame around the code
   keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
   keywordstyle=\color{blue},       % keyword style
   language=C++,                 % the language of the code
   otherkeywords={*,...},           % if you want to add more keywords to the set
   numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
   numbersep=5pt,                   % how far the line-numbers are from the code
   numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
   rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
   showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
   showstringspaces=false,          % underline spaces within strings only
   showtabs=false,                  % show tabs within strings adding particular underscores
   stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
%   stringstyle=\color{mymauve},     % string literal style
   tabsize=2,	                   % sets default tabsize to 2 spaces
   title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}
%\usepackage{layouts}
%\usepackage{mathrsfs}
%\usepackage{pgfkeys,pgfmath,pgfcore}
%\usepackage{srcltx}
%\usepackage[labelformat=simple]{subcaption}
%\usepackage{xspace}
\usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}

@@

\section{Matrix-free operator evaluation}
Classically the approach for finding a discrete solution to a linearized partial differential equation can be described as follows:
First, ansatz spaces and a suitable bilinear form is defined. Then, a matrix and force vector corresponding to the discretization are assembled.
Finally, a direct or iterative solver is used to solve the linear system for the discretized solution.
On modern computer architectures however, it turns out that loading data is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations in the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix vector products. More precisely, the linear system corresponding to a bilinear form 
$a\colon V_h\times V_h\to\mathbb{R}$ and right-hand side $f\colon V_h\to\mathbb{R}$ defined with respect to the ansatz space $V_h$ can be represented by
\begin{align*}
 &&a(v,u) &= f(v) && \forall v\in V_h\\
 \Longleftrightarrow &&a(\varphi_i, u) &= f(\varphi_i) && \forall \varphi_i \\
 \Longleftrightarrow &&\sum_j a(\varphi_i,\varphi_j) u_j &= f(\varphi_i) && \forall \varphi_i \\
 \Longleftrightarrow &&A x &= b
\end{align*}
where $A_{i,j}=a(\varphi_i, \varphi_j)$ and $b_i= f(\varphi_i)$.
Thus, the matrix-vector product $Ax$ can be expressed by
\begin{align*}
 (Ax)_i &= \sum_j a(\varphi_i,\varphi_j) x_j \\
        &= \sum_K\sum_q \sum_j \tilde{a}_K(\varphi_i,\varphi_j)(x_q) x_j w_q 
\end{align*}
where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)$ is a representation for the evaluation of the bilinear in one quadrature point and a quadrature rule is used.
For the sake of demonstration, let us assume $A$ represents a mass matrix. In this case, it holds 
\begin{align*}
 \tilde{a}_K(\varphi_i,\varphi_j)(x_q) = \varphi_i(x_q)\varphi_j(x_q).
\end{align*}
Further assuming that the number of the one-dimensional quadrature points $n$ matches the degree of the ansatz space plus one, the evaluation of all the shape functions ($n^d$) 
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in case we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$
and this is the same as for a regular matrix-vector where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, we can actually do better
than that. Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
\begin{lstlisting}
template<int k, int q>
void evaluate_dof_vector_in_all_quadrature_points
  (const double u[k][k][k],
   const double phi[k][q],
   double return_value[q][q][q])
{
  double tmp1[k][k][q] {};
  for (int i1=0; i1<k; ++i1)
    for (int i2=0; i2<k; ++i2)
      for (int q3=0; q3<q; ++q3)
        for (int i3=0; i3<k; ++i3)
          tmp1[i1][i2][q3] += u[i1][i2][i3]*phi[i3][q3];

  double temp2[k][q][q] {};
  for (int i1=0; i1<k; ++i1)
    for (int q3=0; q3<q; ++q3)
      for (int q2=0; q2<q; ++q2)
        for (int i2=0; i2<k; ++i2)
          tmp2[i1][q2][q3] += tmp1[i1][i2][q3]*phi[i2][q2];

  for (int q2=0; q2<q; ++q2)
    for (int q3=0; q3<q; ++q3)
      for (int q1=0; q1<q; ++q1)
        for (int i1=0; i1<k; ++i1)
  return_value[q1][q2][q3] += tmp2[i1][q2][q3]*phi[i1][q1];
}
\end{lstlisting}
in three space dimensions. In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-base especially for 3D and high polynomial degree.
More information on techniques to increase the computational density can be found in \cite{kronbichler12,vos10}.

@@

TODO: question to answer -- how much to cache? At least $detF ^ {2/3}$. Other options is stress and strain at quadrature points.
Run benchmarks and see.

@@

TODO: add new results from Cook's and some matrix-inclusion (matrix-pore).
Elliptical holes are easier as we don't need heterogeneous operator.
